Characteristic polynomials of p-adic matrices

### 2017-02-01 charpoly/charpoly.tex
Working on the criteria for dchi to be surjective

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
For a matrix $M \in M_n(K)$ we identify the tangent space $V$
at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.
Let $\com(M)$ denote the comatrix of $M$ (when $M \in \GL_n(K)$,
we have $\com(M) = \det(M) M^{-1}$).  Recall
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
that $d\chi(M)$ is given by
\begin{equation} \label{eq:dchi}
d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
\end{equation}
Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$.

@@

\begin{proof}
We first show that the distinct generalized eigenvalues condition
guarantees that $d\chi(M): V \to W$ is surjective.  For any $A \in \GL_n(K)$,
the image of $d\chi(M)$ will be the same as the image of $d\chi(AMA^{-1})$ ,
so we may assume that $M$ is in Jordan form.  Letting $dM$ range over
matrices with a single $1$ entry and all other entries $0$, it suffices to show
that the entries of $\com(X-M)$ span $W$.  Suppose that the blocks of $M$
have dimensions $d_1, \dots, d_m$ with corresponding generalized eigenvalues
$\lambda_1, \dots, \lambda_m$.  The only nonzero entries of $\com(X-M)$
are those on or below the diagonal within the Jordan blocks, and within the $i$th
block, each entry along the $j$th subdiagonal is
$\pm(X-\lambda_i)^{d_i - j - 1} \prod_{k \ne i} (X-\lambda_k)^{d_k}$.
As long as the $\lambda_i$ are distinct, these form a basis for $W$.

@@

\begin{prop}
\label{prop:surjectivity}
For $M \in M_n(K)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item the differential $d\chi(M)$ is surjective
\item the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
to a companion matrix)
\item the eigenspaces of $M$ associated to eigenvalues in an algebraic
closure of $\bar K$ have all dimension $1$
%\item the Jordan blocks of $M$ all have distinct generalized eigenvalues
\end{enumerate}
\end{prop}

@@

\begin{prop}
Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
which $P(M) \in M_n(\OK)$. Then $d \chi_M \big(M_n(\OK)\big)$ 
(considered as a submodule of $K[X]/\chi_M(X)$) is stable by 
multiplication by the elements of $A$.
\end{prop}
\begin{proof}
\todo{The image is generated by the entries of $C = \com(X-M)$
and we have $P(M) \cdot C \equiv  P(X) \cdot C \pmod{\chi_M(X)}$ since the entries of $P(X) \cdot C$
are linear combinations with integer coefficients of the entries of $C$.  $C*(X-M) = \chi(M)$ (since matrix * comatrix is determinant).}
\end{proof}

\begin{cor}
If the coefficients of $M$ lie in $\OK$, then $d \chi_M \big(M_n(\OK)\big)$ 
is stable by multiplication by $X$ and hence is a module over $\OK[X]$.
\end{cor}

\begin{prop}
Let $M \in M_n(K)$, $r>0$ and $H \subset K[X] / \chi(X)$ be the image
of $B_V(r)$ under $d\chi(M)$.  Then $X \cdot H = H$.
\end{prop}
\begin{proof}
\todo{Finish proof}
\end{proof}

>>>>

Let $\chi : M_n(K) \to K_n[X]$ be the characteristic polynomial map.
Our analysis of the precision behavior of $\chi$ rests upon
the computation of its derivative $d\chi$, using \cite{caruso-roe-vaccon:14a}*{Lem. 3.4}.
For a matrix $M \in M_n(K)$ we identify the tangent space $V$
at $M$ with $M_n(K)$ itself, and the tangent space $W$ at $\chi(M)$ with
the space $K_{<n}[X]$ of polynomials of degree less than $n$.
Let $\com(M)$ denote the comatrix of $M$ (when $M \in \GL_n(K)$,
we have $\com(M) = \det(M) M^{-1}$).  Recall
\citelist{\cite{caruso-roe-vaccon:14a}*{Appendix B} \cite{caruso-roe-vaccon:15a}*{\S 3.3}}
that $d\chi(M)$ is given by
\begin{equation} \label{eq:dchi}
d\chi(M): dM \mapsto \tr(\com(X-M) \cdot dM).
\end{equation}

\begin{prop}
\label{prop:surjectivity}
For $M \in M_n(K)$, the following conditions are equivalent:
\begin{enumerate}[(i)]
\renewcommand{\itemsep}{0pt}
\item \label{pt:surj} the differential $d\chi(M)$ is surjective
\item \label{pt:cyc} the matrix $M$ has a cyclic vector (\emph{i.e.} $M$ is similar
to a companion matrix)
\item \label{pt:1d} the eigenspaces of $M$ over the algebraic
closure $\bar{K}$ of $K$ all have dimension $1$
%\item the Jordan blocks of $M$ all have distinct generalized eigenvalues
\end{enumerate}
\end{prop}

\begin{proof}
The equivalence of \eqref{pt:cyc} and \eqref{pt:1d} is standard; see
\cite{hoffman-kunze:LinearAlgebra}*{\S 7.1} for example.  We now show
that \eqref{pt:surj} and \eqref{pt:cyc} are equivalent.

For any $A \in \GL_n(K)$, the image of $d\chi(M)$ will be the same
as the image of $d\chi(AMA^{-1})$, so we may assume that
$M$ is a companion matrix.  For a companion matrix, the bottom row of
$\com(X-M)$ consists of $1, X, X^2, \dots, X^{n-1}$ so $d\chi(M)$ is surjective.

To show the inverse, suppose that $M$ has a repeated eigenvalue $\lambda$ over $\bar{K}$.
After conjugating into Jordan normal form over $\bar{K}$, the entries of $\com(X-M)$
will also be block diagonal, and divisible within each block by the product of $(X-\mu)$
for each other eigenvalue $\mu$.  Since $\lambda$ occurs twice as an eigenvalue,
$X - \lambda$ will divide every entry of $\com(X-M)$ and $d\chi(M)$ will not be surjective.
\end{proof}

Write $B^-_V(r)$ (resp. $B_V(r)$) for the open (resp. closed) ball of radius $r$ in $V$.

@@

\begin{proof}

@@

@@

\begin{prop}
Let $A$ be the subring of $K[X]$ consisting of polynomials $P$ for
which $P(M) \in M_n(\OK)$. Then $d \chi_M \big(M_n(\OK)\big)$ 
(considered as a submodule of $K[X]/\chi_M(X)$) is stable under 
multiplication by the elements of $A$.
\end{prop}
\begin{proof}
\todo{The image is generated by the entries of $C = \com(X-M)$
and we have $P(M) \cdot C \equiv  P(X) \cdot C \pmod{\chi_M(X)}$ since the entries of $P(X) \cdot C$
are linear combinations with integer coefficients of the entries of $C$.  $C*(X-M) = \chi(M)$ (since matrix * comatrix is determinant).}
\end{proof}

\begin{cor}
If the coefficients of $M$ lie in $\OK$, then $d \chi_M \big(M_n(\OK)\big)$ 
is stable under multiplication by $X$ and hence is a module over $\OK[X]$.
\end{cor}
